Boredom-driven curious learning by Homeo-Heterostatic Value Gradients

### 2018-12-16 rebuttal-revision-v2.tex

\usepackage[dvipsnames,table,xcdraw]{xcolor}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{authblk}
\usepackage{float}
\usepackage{tikz}
\usepackage[threshold=2, autopunct=true, autostyle=true]{csquotes}
\usepackage[colorinlistoftodos]{todonotes}

% ============================================================================ %
%                               Macros from Acer                               %
% ============================================================================ %
% reply session
\newenvironment{reply}  
    {\color{Blue}\noindent\newline}
    {\newline}

% \revise
%   {Page}{original text}{revised text}
\newcommand{\revise}[3]{
    \noindent
    \newline
    \textbf{On page {#1}:}\newline
    \newline
    Original:\newline
    \textit{"#2"}
    \newline
    \newline
    Revised:\newline
    \textit{"#3"}\newline}
% ============================================================================ %

\begin{document}
    \section*{Q2:}
        \textit{Is the language, specifically the grammar, of sufficient quality? If no, please specify if the authors should send this manuscript to an expert in English and academic writing.} \newline

    Reviewer 3:
    No. There are many sections that I think would benefit from editing. As just one early example:

    ``This means for a given action-outcome pair the associated reinforcing signal is no longer monotonic. Instead, outcome value varies under reappraisal within according to their relevance to or attainment of goal.''

    The second sentence here has some issue: ``under reappraisal within according to'' is confusing, and ``relevance to or attainment of goal'' should be ``... of a goal''. So, minor things this kind need some improvement.

    \begin{reply}
      % TODO
    \end{reply}

    %--- ----- ------- ----- --- ----- ------- ----- --- ----- ------- ---

    \section*{Q6:}
        \textit{Please comment on the method. Key elements to consider: objective errors or fundamental flaws in the methodology; purpose of new method or technique; appropriateness of context; comprehensive description of procedures; quality of figures and tables.}\newline

        Reviewer 3:
        The authors introduce early the notion that S-R is not the only way to learn - we can be goal directed and engage stimulus-outcome-response learning. This sounds a lot like Robert White's (1959) notion of effectance motivation - that is, we are driven to test effects of our actions on the world. In his piece he cites work from Harlow and Berlyne in animals showing they will work for rewards that satisfy curiosity. I think we don't do a good enough job incorporating some of the older work - and to my reading this work is highly relevant here. The full reference is:

@@

        White, R. W. (1959). Motivation reconsidered: The concept of competence. Psychological review, 66(5), 297-333.

        The description of boredom early on is a fairly narrow one. Boredom is not necessarily tied to knowledge acquisition. I think here the authors are really tapping into the notion of monotony first and perhaps meaning/purpose second. A thing can be monotonous and yet lead to new knowledge (data input for example); similarly, a thing can seem meaningless to an individual but contain ample novel information (e.g., a lecture on string theory to an art historian).

        \begin{reply}
            \textbf{General response:}
            We would first like to express our genuine appreciation towards the reviewer's enthusiasm and knowledge in the field which inspired us a great deal. We have therefore become fully aware that the emergence of boredom does not necessarily induce knowledge acquisition for biological agents. There are other types (e.g., low arousal) of boredom-related behavioural patterns.

            In the current work, we formalised boredom from an information-theoretic standpoint---a focus, albeit limited, that ties formally one aspect of boredom to novelty-seeking and model learning in an agent. However, our work, as pointed out by the reviewer, failed to acknowledge boredom studies from the other side of the spectrum. 
            % We also acknowledge other psychological and physiological factors which co-determine whether biological agents choose to execute information-seeking behaviours during decision making processing.

            % In the current study, we limited our focus on the information-seeking aspect of boredom. This allows us to formalise boredom from an information-theoretical standpoint and 
            % This allow us to precisely define boredom in the mathematical way and evaluate the efficiency of information-seeking behaviours driven by boredom.

            % giving contexts and paving ways

            % We understand that the reviewer concerns about our insufficient literature survey on this topic. We very appreciate the comments and suggestions from the reviewer. We have added more relevant and detailed introduction and discussion on this topic based on the reviewer's comments. Please see our replies as follows. 
        \end{reply}

@@

        In addition, the authors don't do justice to the nuanced literature on boredom and arousal. They selectively cite the high arousal camp (London et al., 1972 - they fail to cite Merrifield \& Danckert, 2014) and ignore the low arousal camp (Barmack, 1939; Geiwitz, 1966; Pattyn et al., 2008; Vogel et al., 2012). The van Tilburg work they cite also consistently shows that self-reports of boredom cast it as a low arousal subjective state. A recent publication examined this debate and claimed the boredom is likely both a high AND low arousal state (subjectively if not physiologically; Danckert et al., 2018 Consc \& Cog). Given that their current model depends on one account of boredom (as high arousal prompting exploration) I think it will be important to at least acknowledge that the boredom literature is not settled on this matter.

        \begin{reply}
            \textbf{Multiple, unsettled accounts of boredom:}
            We appreciate this important suggestion from the reviewer. We have added more references from other camps.

            \revise{\pageref{rev:arousal}}
            % org
                {
                  Psychophysiological studies also demonstrated that boredom plays an active role of information-seeking behaviour. Subjects showing higher levels of reported boredom are accompanied by increased autonomic arousal, such as heart rate and galvanic skin response. These findings are in line with our key notion that boredom intrinsically and actively drives learning behaviours \citep{london1972increase, harris2000correlates}.
                }
            % rev
                {
                  Previous studies have shown that boredom accompanies increases \citep{berlyne1960conflict, london1972increase, harris2000correlates} or decreases \citep{barmack1939definition, geiwitz1966structure, pattyn2008psychophysiological, vogel2012definition, mikulas1993essence} in arousal both in physiological and psychological terms (see \cite{eastwood2012unengaged, fahlman2013development, Merrifield2014, Danckert.2018}). Although the precise account of boredom is still largely unsettled, the high arousal standpoint suggests that boredom may play an active role in the emergence of explorative and information-seeking behaviours.
                  % Even though the potential underline psychophysiological mechanisms are still under debate, the high arousal states elicited by boredom suggest that boredom may play an active role of information-seeking behaviour.
                }
        \end{reply}

        Next, the authors suggest that boredom is associated with creativity. Here again, the research is scant and the authors are not citing it well. Larson (1990) suggested that high ratings of boredom led to the production of lower quality essays - note, the direction here. This doesn't suggest boredom begets creativity, but that it hinders it! Gasper \& Middlewood (2014) induced boredom and suggested it was associated with creativity but their analyses lumped boredom and elation together as promotion focused emotions - clearly not ideal. Finally, Mann \& Cadman (2014) made people bored by having them read the phone book and then completing the alternate uses task - they then pre-selected from the bored group those who ALSO reported daydreaming. The day dreamers were more creative on the alternate uses task. No good control group and no capacity to determine whether daydreaming rather than boredom was the true culprit here. Clearly, the research on this is - well, crap! And the two papers cited are theory pieces that don't do justice to the state of play regarding this potential association. To my mind, boredom is highly unlikely to lead to creativity. Boredom merely signals that we are unengaged. Someone who has already cultivated creative practices may respond to that signal by engaging in a creative activity - but this a far cry from claiming that boredom begets creativity.

        \begin{reply}
            \textbf{Boredom and creativity:}
            Larson's participants lacked the degree of freedom to disengage from the essay-writing, whilst the assessment on one's creativity was based on the quality of essays.
            Creativity in Larson's (1990) terms is largely studies in a narrower domain, namely, essay-writing. It only hinders creativity if the bored individual is instructed not to disengage from the boredom-inducing task.
            Boredom: diversify the outcomes one's actions may bring about. These outcomes are quite often those un-encountered before; therefore, they bears novelty and information to an agent. 
            Stop writing an essay; instead, look at what others are writing
            ``inspiring a search for change and variety'' (Vodanovich & Kass, 1990)

            We thank for the reviewer's suggestion. We agree that, instead of a direct causal relation between boredom and creativity, boredom indirectly increases the probability for an agent to engage in creative activities. We changed our expression of this section and also added the counter-example \citep{larson1990emotions} mentioned by the reviewers in to this section.

            \revise{\pageref{rev:creativity}}
            % org
                {
                  Consistent with our argument, evidence also showed that boredom is associated with increase in creativity \citep{harris2000correlates, schubert1977boredom, schubert1978creativity}. This suggests that the presence of boredom serves to reconfigure agent's instrumental device in order to escape devalued states.
                }
            % rev
                {
                  Based on our notion, we also postulate that boredom may promote agents to perform creative behaviors via disengaging from low information gain environments and re-engaging in creative activities \citep{harris2000correlates, schubert1977boredom, schubert1978creativity}. This suggests that the presence of boredom serves to reconfigure agent's instrumental device in order to escape devalued states (but not in all conditions, see \cite{larson1990emotions}).
                }

@@

        \end{reply}

        Which brings me to another missing piece here: Gomez-Ramirez \& Costa  (2017) (not cited in this paper) published a computational model of boredom and its relation to exploration, with a particular focus on solving the dark room problem. They titled it "Boredom begets creativity" (followed by some subtitle). I like this paper a lot but the title is misleading. Boredom begets the drive for exploration is more appropriate in my view and I think that is what is being proposed here too. But at the very least the authors would need to cite this model and differentiate themselves from it. I think they can do that (the mere notion that their model outdoes other curiosity based agents is enough for me to say this is worth publishing) - but it does need to be done. Otherwise, we're just engaging parallel lines of thinking/research and failing to talk to one another.

        \begin{reply}
            Thank the reviewer for providing this work closely related to our study.
            Ours is about learning to generate behaviours (from an action policy) that can be interpreted as curiosity (here, obtaining a better predictive model of the environment faster).
            Theirs: a very strong prior on how boredom should vary (via the discount factor) over time. An idealised depiction of the interplay between boredom and predictive tendency. Did not address a learning problem. Interesting to frame the problem under the context of financial modelling. Boredom itself does not fit into the picture of predictive coding (would be nice if boredom is also part of the stochastic differential equation).
            Where should I put the text for comparison?

            Computational accounts on boredom are scarce.
            \newline
            \todo{Waiting for Yen}
        \end{reply}

        Just prior to the discussion of Markov Decision processes the authors introduce their notion that boredom is associated with homeostasis and curiosity with heterostasis. I need this to be unpacked more clearly. In my view of boredom it is a drive state arising from a feeling that one's cognitive resources are being underutilised (so the agent is understimulated and needs to do something to raise that). Think of a sustained attention task - these are inherently boring, low rate, low information tasks. As you continue doing it your arousal levels dip perhaps signalling disengagement from the task. You then have two options: do something else or try to raise arousal levels to the necessary point to successfully continue with the task. This sounds like a homeostatic motivation - but is this how the authors are considering things? It wasn't clear. And the flipside was even more opaque.

        \begin{reply}
            We moved more detailed description from session "HOMEO-HETEROSTATIC VALUE GRADIENTS" to this paragraph. We hope that this amendment makes our notion more clear now.

### 2018-12-16 ms.tex
Incorporating yen's revision into the main text.

Animal studies have shown that learning stimulus-response (S-R) association through action-outcome reinforcement is but one facet of instrumental behaviour. Internally, animals may build models that assign values to reappraise experienced outcomes. This expands the landscape of instrumental behaviour to include stimulus-outcome-response (S-O-R) learning system---or goal-directed learning \citep{Balleine.1998}. Goal-directed behaviour is known in both empirical and computational approaches to support adaptive and optimal action selection \citep{adams1981instrumental, adams1982variations, mannella2016goal}. Central to such behavioural adaptiveness is devaluation. This means for a given action-outcome pair the associated reinforcing signal is no longer monotonic. Instead, outcome value varies under reappraisal within according to their relevance to or attainment of goal.

One classic paradigm of devaluation manipulates agent's level of satiation based on food accessibility, leading to altered behavioural pattern. In the context of epistemic disclosure, an analogy can be drawn between devaluation and the emergence of boredom, in which one's assimilation of knowledge reduces the value of similar knowledge in future encounter. The relationship between boredom and outcome devaluation has a long history in psychological research. Empirical findings indicated that boredom is reportedly accompanied by negative affective experience, suggesting that experienced outcomes are intrinsically evaluated and considered as less valuable \citep{bench2013function, van2012boredom, fahlman2009does, perkins1985cognitive, vodanovich1991boredom}. Psychophysiological studies also demonstrated that boredom plays an active role of information-seeking behaviour. Subjects showing higher levels of reported boredom are accompanied by increased autonomic arousal, such as heart rate and galvanic skin response. These findings is in line with our key notion that boredom intrinsically and actively drives agents learning behaviours \citep{london1972increase, harris2000correlates}. Consistent with our argument, evidence also showed that boredom is associated with increase in creativity \citep{harris2000correlates, schubert1977boredom, schubert1978creativity}. This suggests that the presence of boredom serves to reconfigure agent's instrumental device in order to escape devalued states.

Curiosity, irrespective of being a by-product of external goal-attainment or an implicit goal in and of agent itself, is often ascribed as a correlate of information-seeking behaviour \citep{2013gottlieb}. Behaviours exhibiting curious quality are observed in humans and animals alike, suggesting an universal role of curiosity in shaping one's fitness in terms of survival chance. Though the exact neural mechanism underlying the emergence of curious behaviour still remains obscure, current paradigms have their focus on (1) novelty disclosure and (2) uncertainty reduction aspects of information-seeking \citep{pathak2017icm, friston2017curiosity, bellemare2016count, ostrovski2017count}. Indeed, both aspects can be argued to improve agent's fitness in epistemic landscape if the agent elects to incorporate the novelty or uncertainty.

In intrinsic motivation literature \citep{intrinsicmotiv}, although one can readily associate boredom with homeostatic motivation and curiosity with heterostatic motivation, our argument suggests they can in fact be complementary. Our contribution thus pertains to the reconciliation of homeo-heterostatic motivations.

@@

The algorithm offers reconciliation between the two seemingly opposite qualities and concludes with their cooperative nature. Specifically, the knowledge an organism maintains about its comfort zone helps instigate outbound heterostatic drives. In return, satisfying heterostatic drives broadens the organism's extent of comfort zone. As a consequence, the organism not only improves its fitness in terms of homeostatic outreach but also becomes effectively curious.

@@

One rewarding scenario happens when $Q(S^\prime|\bs{s}; \psi)$ has all the information there is to be possessed by $A$ about $S^\prime$. $A$ is therefore rendered redundant. One may speculate, at this point, the agent could opt for inhibiting its responses. Disengaging actions potentially saves energy which is rewarding in biological sense.

@@

\citet{intrinsicmotiv} outlined the distinctions between two important classes of intrinsic motivation: homeostatic and heterostatic. A homeostatic motivation is one that can be satiated, leading to certain equilibrium behaviourally; whereas a heterostatic motivation topples the agent, thus preventing it from occupying habitual states.

@@

We have provided a formal account on the emergence of boredom from an information-seeking perspective and addressed its constructive role in enabling curious behaviour. Boredom thus motivated an instrumental view of action selection, in which an action serves to disclose outcomes that have intrinsic meaningfulness to an agent itself. This is, a bored agent must seek out information worth assimilating into itself. This led to the central claim of this study---pertaining to superior data-gathering efficiency and hence effective curiosity. We supported this claim with empirical evidence, showing that boredom-enabled agents consistently outperformed other curious agents in self-assisted forward model learning. Our results solicited the interpretation that the relationship between homeostatic and heterostatic intrinsic motivations can in fact be complementary; therefore, we have offered one unifying perspective for the intrinsic motivation landscape.

>>>>

Animal studies have shown that learning stimulus-response (S-R) associations through action-outcome reinforcement is but one facet of instrumental behaviour. Internally, animals may build models that assign values to reappraise experienced outcomes. This expands the landscape of instrumental behaviour to include the stimulus-outcome-response (S-O-R) learning system---or goal-directed learning \citep{Balleine.1998}. Goal-directed behaviour is known in both empirical and computational approaches to support adaptive and optimal action selection \citep{adams1981instrumental, adams1982variations, mannella2016goal}. Central to such behavioural adaptiveness is devaluation. This means for a given action-outcome pair the associated reinforcing signal is no longer monotonic. Instead, an outcome value will change with reappraisals in accordance with an agent's internal goal.

One classic paradigm of devaluation manipulates an agent's level of satiation based on food accessibility, leading to altered behavioural patterns. In the context of epistemic disclosure, an analogy can be drawn between devaluation and the emergence of boredom, in which one's assimilation of knowledge reduces the value of similar knowledge in future encounters. The relationship between boredom and outcome devaluation has a long history in psychological research. Empirical findings indicated that boredom is reportedly accompanied by negative affective experiences, suggesting that experienced outcomes are intrinsically evaluated and considered as less valuable \citep{bench2013function, van2012boredom, fahlman2009does, perkins1985cognitive, vodanovich1991boredom}.

\label{rev:arousal}
Psychophysiological studies also demonstrated that boredom plays an active role in eliciting information-seeking behaviours. Subjects showing higher levels of reported boredom are accompanied by increased autonomic arousal, such as heart rate and galvanic skin response. These findings are in line with our key notion that boredom intrinsically and actively drives learning behaviours \citep{berlyne1960conflict, london1972increase, harris2000correlates}. Note, however, that this notion is contested and a matter of unsettled debate (e.g., \citealp{eastwood2012unengaged, fahlman2013development, Merrifield2014, Danckert.2018}). It is therefore worth pointing out that boredom may be accompanied by a low arousal state \citep{barmack1939definition, geiwitz1966structure, pattyn2008psychophysiological, vogel2012definition, mikulas1993essence}.

A finding by Larson \citeyearpar{larson1990emotions} invites the speculation that a task set may interact with boredom, thereby modifying a subject’s behavioural pattern to follow either low or high arousal states. This means boredom may merely signal a state of disengagement. Whether an agent’s cognitive resources can be freely allocated to re-engage another task inherently depends upon the existence of a prohibiting condition. Larson’s \citeyearpar{larson1990emotions} participants, who reported boredom and were later rated with low scores in creative writing, were by design not allowed to disengage from the essay-writing task. Other theories, on the other hand, suggested that boredom is associated with increase in creativity \citep{harris2000correlates, schubert1977boredom, schubert1978creativity}.

We thus postulate that, in the absence of any {\it a priori} cognitive or behavioural constraints, a state of boredom is followed by an attempt to diversify one’s experience. That is, boredom begets exploration. This is in line with Vodanovich and Kass’s \citeyearpar{vodanovich1990factor} notion of boredom in ``inspiring a search for change and variety'' and Zuckerman’s \citeyearpar{zuckerman2008sensation} ``sensation-seeking''. Sensation-seeking \citep{zuckerman1971dimensions, zuckerman2008sensation, dahlen2005driving, kass1990boredom} is categorised as a personality trait, tightly linked to boredom susceptibility \citep{zuckerman1978sensation}. High sensation seekers get bored more easily, suggesting that individuals susceptible to boredom are predisposed to seek novel sensations. As a result, a learner who is also a novelty-seeker may have a world model that generalises better. In our framework, receiving novel sensations is formalised as planning to visit states where an agent can effectively learn faster (i.e., the agent gets bored quicker). This effect is then treated as an intrinsic reward, prompting an agent to continue experiencing the state before the reward is depleted.

\label{rev:creativity}
A recent computational modelling tapped into a similar theme \citep{Gomez-Ramirez2017}, where boredom facilitates exploration. However, our work differs from that of Gomez-Ramirez and Costa \citeyearpar{Gomez-Ramirez2017} in that our model permits a simple form of agency (by having an action policy) and focuses on learning. Additionally, their exploration may favour predictable state space, whereas our agent will treat high predictability as an intrinsically non-rewarding state.

Finally, in psychology studies, the term boredom usually comes under two distinct constructs: a state of boredom and boredom proneness \citep{Elpidorou2014, Elpidorou.2017, mugon2018failure}. Boredom proneness is regarded as the psychological predisposition of an individual to experience boredom which poses a systematic impact on one’s social and psychological well-being. By contrast, a state of boredom is seen as a transient, regulatory signal that prompts one’s behaviours into alignment with its goal-directedness \citep{Elpidorou.2017}. In this sense, our model conceptually encompasses the function of the state boredom regulatory signal.

Curiosity, irrespective of being a by-product of external goal-attainment or an implicit goal in and of an agent itself, is often ascribed as a correlate of information-seeking behaviour \citep{2013gottlieb}. Behaviours exhibiting curious quality are observed in humans and animals alike, suggesting an universal role of curiosity in shaping one's fitness in terms of survival chance. Though the exact neural mechanism underlying the emergence of curious behaviour still remains obscure, current paradigms have their focus on (1) novelty disclosure and (2) uncertainty reduction aspects of information-seeking \citep{pathak2017icm, friston2017curiosity, bellemare2016count, ostrovski2017count}. Indeed, both aspects can be argued to improve agent's fitness in epistemic landscape if the agent elects to incorporate the novelty or uncertainty.

Both boredom and curiosity are tightly connected to the notion of intrinsic motivation. Specifically, the occurrence of boredom and curiosity can be mapped to homeostatic and heterostatic motivations, respectively. The homeostatic and heterostatic motivations as two important classes of intrinsic motivation have been extensively reviewed in Oudeyer and Kaplan \citeyearpar{intrinsicmotiv}. Simply, a homeostatic motivation drives a system to compensate perturbations in order to reach some equilibrial state. A heterostatic motivation is the opposite of a homeostatic motivation. A system that is driven by heterostatic motivations will self-perturb out of its equilibrium. In our formalism, predictive model learning and policy learning, each respectively induces boredom and curiosity, suggesting that the two classes of motivation can in fact be complementary when the two learning tasks are carried out concurrently. Our contribution thus pertains to the reconciliation of homeo-heterostatic motivations.

@@

The algorithm offers a reconciliation between the two seemingly opposite qualities and concludes with their cooperative nature. Specifically, the knowledge an organism maintains about its comfort zone helps instigate outbound heterostatic drives. In return, satisfying heterostatic drives broadens the organism's extent of comfort zone. As a consequence, the organism not only improves its fitness in terms of homeostatic outreach but also becomes effectively curious.

@@

One rewarding scenario happens when $Q(S^\prime|\bs{s}; \psi)$ has all the information there is to be possessed by $A$ about $S^\prime$. $A$ is therefore rendered redundant. One may speculate, at this point, the agent could opt for inhibiting its responses. Disengaging actions potentially saves energy which is rewarding in biological sense. \label{rev:energyPOV} This outcome is in line with the ``opportunity cost model'' proposed by Kurzban et al. \citeyearpar{Kurzban2013}. In their model, boredom is seen as a resource regulatory signal which drives an agent to disengage the current task and curb the computational cost. As a consequence, the occurrence of boredom may encourage re-allocation of computational processes to alternative higher-value activities \citep{Kurzban2013}.

@@

\citet{intrinsicmotiv} outlined the distinctions between two important classes of intrinsic motivation: homeostatic and heterostatic. A homeostatic motivation is one that can be satiated, leading to a certain equilibrium behaviourally; whereas a heterostatic motivation topples the agent, thus preventing it from occupying habitual states.

@@

We have provided a formal account on the emergence of boredom from an information-seeking perspective and addressed its constructive role in enabling curious behaviour. Boredom thus motivated an instrumental view of action selection, in which an action serves to disclose outcomes that have intrinsic meaningfulness to an agent itself. This is, a bored agent must seek out information worth assimilating into itself. This led to the central claim of this study---pertaining to the superior data-gathering efficiency and hence effective curiosity. We supported this claim with empirical evidence, showing that boredom-enabled agents consistently outperformed other curious agents in self-assisted forward model learning. Our results solicited the interpretation that the relationship between homeostatic and heterostatic intrinsic motivations can in fact be complementary; therefore, we have offered one unifying perspective for the intrinsic motivation landscape.
